Robust Bayesian Analysis of Causal Inference in Medical Diagnosis and Treatment

### 2024-08-02 JASA-template.tex
Thank reviewers.
Respond to confusing terminology.

%%% MT to TB: what about "Robust Bayesian Analysis of Causal Inference in Medical Diagnosis and Treatment"
%%% TB to MT: changed it
\title{Robust Bayesian Analysis of Causal Inference in Medical Diagnosis and Treatment}
\author[1]{Tathagata Basu\corref{cor1}}
\ead{tathagatabasumaths@gmail.com}
\cortext[cor1]{Corresponding author}
\author[2]{Matthias C.~M.~Troffaes}
\ead{matthias.troffaes@durham.ac.uk}
\affiliation[1]{organization={Civil and Environmental Engineering, University of Strathclyde},
            addressline={16 Richmond St}, 
            city={Glasgow},
            postcode={G1 1XQ}, 
            %state={},
            country={United Kingdom}}
\affiliation[2]{organization={Department of Mathematical Sciences, Durham University},
            addressline={South Road}, 
            city={Durham},
            postcode={DH13LE}, 
            %state={},
            country={United Kingdom}}

@@

\begin{highlights}
\item Earlier a version of this paper was presented in ECSQARU 2023 conference. 
Since then we have changed the following contents of the paper for this extended version. 
\item A thorough introduction with additional image to express our problem
formulation in a better way as well as discussion of previous works.
\item Introduction of additional equations to justify and explain the
causal inference model under confounding.
\item We have put additional text for better explanation of Bayesian causal
estimation model as well as the importance of a robust Bayesian analysis. These
are accompanied by a new subsection discussing refit model, which can be
used for better understanding and selection of a single model from set of causal models.
\item A larger simulation study with more number of observations. Additionally
we have introduced a new simulation setting where we vary the number of predictors
for better understanding of the performance of our approach.
\end{highlights}

@@

The hierarchical model presented above is a standard spike and slab model for
variable selection and performs well when we have sufficient data. 
However, especially in the case of causal inference, we do not always have sufficient data. Moreover, 
we also must be cautious about the side effects of a treatment.
Therefore, we are particularly interested in constructing a robust Bayesian framework
for variable selection. In this way, when we are preparing guidelines for treatment, we 
can have the option to ask for more data before reaching any conclusion. To achieve this,
we consider a utility based framework with three possible 
scenarios.

@@

For the simulation studies, we consider 4 different settings. In each
case, we generate the design matrix $X$ such that $X_i\sim\mathcal{N}(0, \Sigma)$
for $1\le i\le n$ where $\Sigma_{ij} = 0.3^{|i-j|}$. In this way, we 
generate predictors for our model with mild correlations between them.
We then use the following distributions to generate the outcome and
the treatment indicator: 
\begin{equation}
    T_i \sim \text{Bernoulli}\left(1/(1+\exp(-X_i\gamma))\right)
    \quad\text{and}\quad
    Y_i = 4T_i + X_i\beta + \epsilon_i.
\end{equation}
where $\epsilon_i\sim\mathcal{N}(0,0.1^2)$.
Note that the simulated causal effect $\beta_T$ is equal to $4$.

These four cases can be split in two categories. In one category we consider the
case where we have increasing number of observations. In the first category we have two
sub cases, in the first case we consider all active variables to be confounders
and in the second case we consider some active variables which are only related to
the outcome model. 
\begin{description}
    \item[Scenario 1a] --- $|\gamma_j|, |\beta_j|>0$ for $j\le 10$
    \item[Scenario 1b] --- $|\gamma_j|>0$ for $j\le 10$ and $|\beta_j|>0$ for $j\le 15$
\end{description}
For both
scenarios, we consider different numbers of observations $n$ where
$n=25+ 5k$ for $k=0,1,2,\dots,25$ and $p=50$
predictors. This way, we check the efficiency of our method with varying level of information.

For the second category, we check our method for varying number of predictors 
(and hence sparsity level, i.e.\ the percentage of active variables present in the model).
Similar to the first category we also have two sub cases:
\begin{description}
    \item[Scenario 2a] --- $|\gamma_j|, |\beta_j|>0$ for $j\le 10$
    \item[Scenario 2b] --- $|\gamma_j|>0$ for $j\le 10$ and $|\beta_j|>0$ for $j\le 15$
\end{description}
For both
cases, we consider different numbers of predictors $p$ where
$p=25+ 5k$ for $k=1,2,\dots,25$ and $n=100$
subjects.

@@

\paragraph{Results}
\Cref{tab:causal1} shows the results of estimating the causal effect $\beta_T$
for the first and second scenarios.
For reference, recall the true value is $\beta_T=4$. 
As we perform a sensitivity analysis,
our method gives an interval estimate for the causal effect:
\added{column RBCE-l gives the lower bound
and column RBCE-u gives the upper bound}. We notice that our method is 
somewhat in agreement with the other methods but is more consistent
in terms of estimating the causal effect in the first setting. However, this is not the
case for the analysis with second dataset. We notice that our method tends to under
estimate the causal effect. However, sometimes, other methods produce extreme 
values which is not the case for our method. This can be observed in \cref{fig:comp:trt} as well. %Here, the true 
%value is represented by the black straight line for $\beta_T = 4$.

@@

% latex table generated in R 4.3.0 by xtable 1.8-4 package
% Tue May 23 18:29:20 2023
% latex table generated in R 4.3.2 by xtable 1.8-4 package
% Mon Mar 25 09:52:08 2024
\begin{table}[ht]
\centering
\tiny
\caption{Comparison of different methods in estimating the causal effect for varying number of observations.}
\begin{tabular}{c|ccccc|ccccc}
  \hline
 Case&  &  & 1 &  &  &  &  & 2 &  &  \\ 
  \hline
 & RBCE-l & RBCE-u & SSCE & BSSCE & BSSL & RBCE-l & RBCE-u & SSCE & BSSCE & BSSL \\ 
  \hline
25 & 2.61 & 3.38 &  &  & 2.19 & 2.70 & 3.29 &  &  & 8.80 \\ 
  30 & 3.14 & 3.56 &  &  & 3.99 & 3.01 & 3.54 &  &  & 9.02 \\ 
  35 & 3.15 & 3.52 &  &  & 3.98 & 3.22 & 3.66 &  &  & 6.66 \\ 
  40 & 3.60 & 3.75 &  &  & 3.98 & 3.20 & 3.70 &  &  & 4.42 \\ 
  45 & 3.76 & 3.82 &  &  & 3.96 & 3.61 & 3.74 &  &  & 3.99 \\ 
  50 & 3.79 & 3.84 & 3.99 & 4.01 & 3.99 & 3.75 & 3.79 & 9.04 & 9.36 & 4.01  \\ 
  55 & 3.89 & 3.91 & 3.97 & 4.00 & 3.98 & 3.77 & 3.81 & 5.92 & 6.38 & 4.02 \\ 
  60 & 3.93 & 3.95 & 3.99 & 4.00 & 3.98 & 3.76 & 3.77 & 3.89 & 3.50 & 4.03 \\ 
  65 & 3.95 & 3.96 & 4.00 & 4.02 & 4.00 & 3.76 & 3.78 & 3.91 & 3.99 & 4.02 \\ 
  70 & 3.94 & 3.96 & 4.01 & 4.05 & 4.00 & 3.77 & 3.79 & 4.01 & 3.96 & 4.03 \\ 
  75 & 3.97 & 3.98 & 4.02 & 4.03 & 4.02  & 3.82 & 3.83 & 4.03 & 4.02 & 4.04 \\ 
  80 & 3.96 & 3.98 & 4.02 & 4.05 & 4.01 & 3.83 & 3.85 & 4.04 & 4.05 & 4.05 \\ 
  85 & 3.96 & 3.98 & 4.03 & 4.04 & 4.03 & 3.84 & 3.86 & 4.05 & 4.06 & 4.06 \\ 
  90 & 3.97 & 3.99 & 4.05 & 4.07 & 4.04 & 3.87 & 3.88 & 4.06 & 4.04 & 4.05 \\ 
  95 & 3.96 & 3.98 & 4.02 & 4.05 & 4.04 & 3.88 & 3.89 & 4.05 & 4.06 & 4.06 \\ 
  100 & 3.95 & 3.98 & 4.03 & 4.05 & 4.04 & 3.88 & 3.89 & 4.07 & 4.06 & 4.07 \\ 
  105 & 3.97 & 3.98 & 4.03 & 4.05 & 4.03 & 3.89 & 3.90 & 4.07 & 4.06 & 4.07 \\ 
  110 & 3.97 & 3.99 & 4.03 & 4.05 & 4.03 & 3.90 & 3.91 & 4.07 & 4.05 & 4.07 \\ 
  115 & 3.98 & 3.99 & 4.04 & 4.05 & 4.04 & 3.91 & 3.92 & 4.06 & 4.06 & 4.06 \\ 
  120 & 3.98 & 3.99 & 4.04 & 4.05 & 4.03 & 3.91 & 3.92 & 4.04 & 4.05 & 4.04 \\ 
  125 & 3.99 & 3.99 & 4.03 & 4.04 & 4.03 & 3.91 & 3.92 & 4.04 & 4.04 & 4.04 \\ 
  130 & 3.98 & 3.99 & 4.02 & 4.04 & 4.02 & 3.92 & 3.93 & 4.03 & 4.04 & 4.04 \\ 
  135 & 3.99 & 3.99 & 4.02 & 4.04 & 4.02 & 3.94 & 3.94 & 4.04 & 4.05 & 4.04 \\ 
  140 &  3.99 & 4.00 & 4.03 & 4.05 & 4.03 & 3.94 & 3.95 & 4.04 & 4.05 & 4.04 \\ 
  145 & 3.99 & 3.99 & 4.03 & 4.04 & 4.03 & 3.95 & 3.95 & 4.05 & 4.05 & 4.05 \\ 
  150 & 3.99 & 3.99 & 4.02 & 4.02 & 4.02 & 3.94 & 3.95 & 4.04 & 4.04 & 4.04 \\ 
   \hline
\end{tabular}
\label{tab:causal1}
\end{table}
\begin{figure}
    \centering
    \includegraphics[width = 0.9\linewidth]{treat_obs.pdf}
    \caption{Comparison of different methods in estimating the causal effect for varying number of observations.}
    \label{fig:comp:trt}
\end{figure}

@@

From the table it can be seen that for the first scenario, our method

@@

for the first setting and second setting respectively. However,

@@

\caption{Loss based on misspecification of active variables for first scenario.}\label{tab:misspec1}

@@

\caption{Loss based on misspecification of active variables for second scenario}

@@

Besides this, we notice that all of the four methods perform
perfectly in variable selection giving no misspecified variables this
can also be understood from \cref{fig:comp:trt:pred} because
in case of misspecified variables we get a very different value
of the treatment effect. Therefore,
we omit the table of variable selection for third and fourth scenarios.

@@

We sincerely thank Jochen Einbeck for his contributions to an earlier version of this paper.

>>>>

%%% MT what about using "Robust Bayesian causal estimation" (name for our method) in the title, e.g. "Robust Bayesian causal estimation for causal inference in medical diagnosis"
\title{Robust Bayesian Analysis of Causal Inference in Medical Diagnosis and Treatment}
\author[1]{Tathagata Basu\corref{cor1}}
\ead{tathagatabasumaths@gmail.com}
\cortext[cor1]{Corresponding author}
\author[2]{Matthias C.~M.~Troffaes}
\ead{matthias.troffaes@durham.ac.uk}
\affiliation[1]{organization={Civil and Environmental Engineering, University of Strathclyde},
            addressline={16 Richmond St}, 
            city={Glasgow},
            postcode={G1 1XQ}, 
            %state={},
            country={United Kingdom}}
\affiliation[2]{organization={Department of Mathematical Sciences, Durham University},
            addressline={South Road}, 
            city={Durham},
            postcode={DH13LE}, 
            %state={},
            country={United Kingdom}}

@@

\begin{highlights}
\item Earlier a version of this paper was presented in ECSQARU 2023 conference. 
Since then we have changed the following contents of the paper for this extended version. 
\item A thorough introduction with additional image to express our problem
formulation in a better way as well as discussion of previous works.
\item Introduction of additional equations to justify and explain the
causal inference model under confounding.
\item We have put additional text for better explanation of Bayesian causal
estimation model as well as the importance of a robust Bayesian analysis. These
are accompanied by a new subsection discussing refit model, which can be
used for better understanding and selection of a single model from set of causal models.
\item A larger simulation study with more number of observations. Additionally
we have introduced a new simulation where we vary the number of predictors
for better understanding of the performance of our approach.
\end{highlights}

@@

The hierarchical model presented above is a standard spike and slab model for
variable selection and performs well when we have sufficient data. 
However, especially in the case of causal inference, we do not always have sufficient data. Moreover, 
we also must be cautious about the side effects of a treatment.
Therefore, we are particularly interested in constructing a robust Bayesian framework
for variable selection. In this way, when we are preparing guidelines for treatment, we 
can have the option to ask for more data before reaching any conclusion. To achieve this,
we consider a utility based framework with three possible \added{outcomes}.

@@

For the simulation studies, we consider \added{2 different cases each with 2 sub-cases, amounting to 4 studies in total}. In each
\alert{of these 4 studies}, we generate the design matrix $X$ such that $X_i\sim\mathcal{N}(0, \Sigma)$
for $1\le i\le n$ where $\Sigma_{ij} = 0.3^{|i-j|}$. In this way, we 
generate predictors for our model with mild correlations between them.
We then use the following distributions to generate the outcome and
the treatment indicator: 
\begin{equation}
    T_i \sim \text{Bernoulli}\left(1/(1+\exp(-X_i\gamma))\right)
    \quad\text{and}\quad
    Y_i = 4T_i + X_i\beta + \epsilon_i.
\end{equation}
where $\epsilon_i\sim\mathcal{N}(0,0.1^2)$.
Note that the simulated causal effect $\beta_T$ is equal to $4$.

\added{In case 1, we consider an} increasing number of observations. \added{We} have two
sub-cases: in \alert{case 1a} we consider all active variables to be confounders
and in the second \alert{case 1b} we consider some active variables which are only related to
the outcome model. 
\begin{description}
    \item[\added{Case} 1a] --- $|\gamma_j|, |\beta_j|>0$ for $j\le 10$
    \item[\added{Case} 1b] --- $|\gamma_j|>0$ for $j\le 10$ and $|\beta_j|>0$ for $j\le 15$
\end{description}
For both \added{case 1a and 1b}, we consider different numbers of observations $n$ where
$n=25+ 5k$ for $k=0,1,2,\dots,25$ and $p=50$
predictors. This way, we check the efficiency of our method with varying level of information.

For \added{case 2}, we check our method for varying number of predictors 
(and hence sparsity level, i.e.\ the percentage of active variables present in the model).
Similar to \added{case 1} we also have two \added{sub-cases}:
\begin{description}
    \item[\added{Case} 2a] --- $|\gamma_j|, |\beta_j|>0$ for $j\le 10$
    \item[\added{Case} 2b] --- $|\gamma_j|>0$ for $j\le 10$ and $|\beta_j|>0$ for $j\le 15$
\end{description}
For both
\added{case 2a and 2b}, we consider different numbers of predictors $p$ where
$p=25+ 5k$ for $k=1,2,\dots,25$ and $n=100$
subjects.

@@

\paragraph{Results}
\Cref{tab:causal1} shows the results of estimating the causal effect $\beta_T$
for \added{case 1a and 1b}.
For reference, recall the true value is $\beta_T=4$. 
As we perform a sensitivity analysis,
our method gives an interval estimate for the causal effect:
\added{column RBCE-l gives the lower bound
and column RBCE-u gives the upper bound}. We notice that our method is 
somewhat in agreement with the other methods but is more consistent
in terms of estimating the causal effect in \added{case 1a}. However, this is not the
case for the analysis with \added{case 1b}. We notice that our method tends to under
estimate the causal effect. However, sometimes, other methods produce extreme 
values which is not the case for our method. This can be observed in \cref{fig:comp:trt} as well. %Here, the true 
%value is represented by the black straight line for $\beta_T = 4$.

@@

% latex table generated in R 4.3.0 by xtable 1.8-4 package
% Tue May 23 18:29:20 2023
% latex table generated in R 4.3.2 by xtable 1.8-4 package
% Mon Mar 25 09:52:08 2024
\begin{table}
\centering
\tiny
\caption{Comparison of different methods in estimating the causal effect for varying number of observations.}
\begin{tabular}{c|ccccc|ccccc}
  \hline
 Case&  &  & \added{1a} &  &  &  &  & \added{1b} &  &  \\ 
  \hline
\added{Obs} & RBCE-l & RBCE-u & SSCE & BSSCE & BSSL & RBCE-l & RBCE-u & SSCE & BSSCE & BSSL \\ 
  \hline
25 & 2.61 & 3.38 &  &  & 2.19 & 2.70 & 3.29 &  &  & 8.80 \\ 
  30 & 3.14 & 3.56 &  &  & 3.99 & 3.01 & 3.54 &  &  & 9.02 \\ 
  35 & 3.15 & 3.52 &  &  & 3.98 & 3.22 & 3.66 &  &  & 6.66 \\ 
  40 & 3.60 & 3.75 &  &  & 3.98 & 3.20 & 3.70 &  &  & 4.42 \\ 
  45 & 3.76 & 3.82 &  &  & 3.96 & 3.61 & 3.74 &  &  & 3.99 \\ 
  50 & 3.79 & 3.84 & 3.99 & 4.01 & 3.99 & 3.75 & 3.79 & 9.04 & 9.36 & 4.01  \\ 
  55 & 3.89 & 3.91 & 3.97 & 4.00 & 3.98 & 3.77 & 3.81 & 5.92 & 6.38 & 4.02 \\ 
  60 & 3.93 & 3.95 & 3.99 & 4.00 & 3.98 & 3.76 & 3.77 & 3.89 & 3.50 & 4.03 \\ 
  65 & 3.95 & 3.96 & 4.00 & 4.02 & 4.00 & 3.76 & 3.78 & 3.91 & 3.99 & 4.02 \\ 
  70 & 3.94 & 3.96 & 4.01 & 4.05 & 4.00 & 3.77 & 3.79 & 4.01 & 3.96 & 4.03 \\ 
  75 & 3.97 & 3.98 & 4.02 & 4.03 & 4.02  & 3.82 & 3.83 & 4.03 & 4.02 & 4.04 \\ 
  80 & 3.96 & 3.98 & 4.02 & 4.05 & 4.01 & 3.83 & 3.85 & 4.04 & 4.05 & 4.05 \\ 
  85 & 3.96 & 3.98 & 4.03 & 4.04 & 4.03 & 3.84 & 3.86 & 4.05 & 4.06 & 4.06 \\ 
  90 & 3.97 & 3.99 & 4.05 & 4.07 & 4.04 & 3.87 & 3.88 & 4.06 & 4.04 & 4.05 \\ 
  95 & 3.96 & 3.98 & 4.02 & 4.05 & 4.04 & 3.88 & 3.89 & 4.05 & 4.06 & 4.06 \\ 
  100 & 3.95 & 3.98 & 4.03 & 4.05 & 4.04 & 3.88 & 3.89 & 4.07 & 4.06 & 4.07 \\ 
  105 & 3.97 & 3.98 & 4.03 & 4.05 & 4.03 & 3.89 & 3.90 & 4.07 & 4.06 & 4.07 \\ 
  110 & 3.97 & 3.99 & 4.03 & 4.05 & 4.03 & 3.90 & 3.91 & 4.07 & 4.05 & 4.07 \\ 
  115 & 3.98 & 3.99 & 4.04 & 4.05 & 4.04 & 3.91 & 3.92 & 4.06 & 4.06 & 4.06 \\ 
  120 & 3.98 & 3.99 & 4.04 & 4.05 & 4.03 & 3.91 & 3.92 & 4.04 & 4.05 & 4.04 \\ 
  125 & 3.99 & 3.99 & 4.03 & 4.04 & 4.03 & 3.91 & 3.92 & 4.04 & 4.04 & 4.04 \\ 
  130 & 3.98 & 3.99 & 4.02 & 4.04 & 4.02 & 3.92 & 3.93 & 4.03 & 4.04 & 4.04 \\ 
  135 & 3.99 & 3.99 & 4.02 & 4.04 & 4.02 & 3.94 & 3.94 & 4.04 & 4.05 & 4.04 \\ 
  140 &  3.99 & 4.00 & 4.03 & 4.05 & 4.03 & 3.94 & 3.95 & 4.04 & 4.05 & 4.04 \\ 
  145 & 3.99 & 3.99 & 4.03 & 4.04 & 4.03 & 3.95 & 3.95 & 4.05 & 4.05 & 4.05 \\ 
  150 & 3.99 & 3.99 & 4.02 & 4.02 & 4.02 & 3.94 & 3.95 & 4.04 & 4.04 & 4.04 \\ 
   \hline
\end{tabular}
\label{tab:causal1}
\end{table}
\begin{figure}
    \centering
    \includegraphics[width = 0.9\linewidth]{treat_obs.pdf}
    \caption{Comparison of different methods in estimating the causal effect for varying number of observations.}
    \label{fig:comp:trt}
\end{figure}

@@

From the table it can be seen that for \added{case 1}, our method

@@

for \added{cases 1a and 1b} respectively. However,

@@

\caption{Loss based on misspecification of active variables for \added{case 1a}.}\label{tab:misspec1}

@@

\caption{Loss based on misspecification of active variables for \added{case 1b}}

@@

Besides this, we notice that all of the four methods perform
perfectly in variable selection giving no misspecified variables this
can also be understood from \cref{fig:comp:trt:pred} because
in case of misspecified variables we get a very different value
of the treatment effect. Therefore,
we omit the table of variable selection for \added{cases 2a and 2b}.

@@

We sincerely thank Jochen Einbeck for his contributions to an earlier version of this paper.
We also thank all reviewers for their kind and constructive comments which helped improve the paper.
